---
title: "Cross-Language Numerical Equivalence Validation: Verifying the R Implementation of PairwiseAUCLoss against Python libauc"
author: "Aman Kashyap"
date: "2026-06-25"
categories: [R, torch, benchmark, gsoc]
---

We implemented
[PairwiseAUCLoss](https://github.com/Optimization-AI/LibAUC/blob/main/libauc/losses/auc.py#L855)
from the Python
[libauc](https://docs.libauc.org/api/libauc.losses.html#libauc.losses.auc.PairwiseAUCLoss)
library in R as part of the
[mlr3torchAUM](https://github.com/mlr3-imbalanced/mlr3torchAUM)
package. To guarantee the correctness of our R implementation, we perform **cross-language numerical differential testing** using the original Python library as a reference oracle.

Specifically, we generate predictions and labels in PyTorch, evaluate the loss using both implementations across various input sizes, and verify output equivalence (up to floating-point precision) using the
[atime](https://github.com/tdhock/atime)
R package.

## Setup

First, we load the necessary libraries and ensure the Python
dependencies are installed. Because R's `torch` and Python's `torch`
both ship their own C++ LibTorch backend, loading them in the same
process causes symbol collisions. We therefore use `reticulate` to
cleanly manage our Python virtual environment, but we will run all
Python code in a **separate subprocess** via `system2()` to keep the
C++ runtimes completely isolated.

```{r setup, warning=FALSE, message=FALSE}
library(reticulate)
library(torch)
library(atime)
library(data.table)

# Set the index URL for CPU-only PyTorch so reticulate's `uv` backend 
# fetches the small CPU wheels instead of massive CUDA ones.
Sys.setenv(UV_INDEX = "https://download.pytorch.org/whl/cpu")

# Create isolated environment and install packages
reticulate::virtualenv_create("libauc_env")
reticulate::virtualenv_install("libauc_env", "torch")
reticulate::virtualenv_install("libauc_env", "libauc")

# Get the path to the isolated Python binary to use in subprocesses
py_bin <- reticulate::virtualenv_python("libauc_env")

# Install the R torch C++ backend (Lantern/LibTorch) if not yet downloaded.
if (!torch::torch_is_installed()) {
  torch::install_torch()
}
```

## Python helper script

We write a small standalone Python script that computes the
PairwiseAUCLoss for a given surrogate. It reads predictions and labels
from a temporary CSV, computes the loss, and prints the scalar result
to stdout. Running it as a subprocess avoids loading Python's
libtorch into R's address space.

```{python, eval=FALSE}
import sys, csv, torch
from libauc.losses import PairwiseAUCLoss

surr_name = sys.argv[1]
csv_path  = sys.argv[2]

surrogates = {
    "squared":       {"hparam": 1.0},
    "squared_hinge": {"hparam": 1.0},
    "hinge":         {"hparam": 1.0},
    "logistic":      {"hparam": 2.0},
    "barrier_hinge": {"hparam": (1.0, 2.0)},
}

with open(csv_path) as f:
    reader = csv.reader(f)
    rows = list(reader)

y_pred = torch.tensor([float(x) for x in rows[0]], dtype=torch.float32)
y_true = torch.tensor([float(x) for x in rows[1]], dtype=torch.float32)

cfg = surrogates[surr_name]
loss_fn = PairwiseAUCLoss(surr_loss=surr_name, hparam=cfg['hparam'])
result = loss_fn(y_pred.view(-1, 1), y_true.view(-1, 1)).item()
print(f'{result:.15g}')
```

```{r, echo=FALSE}
py_loss_script <- tempfile(fileext = ".py")
writeLines(con = py_loss_script, c(
  "import sys, csv, torch",
  "from libauc.losses import PairwiseAUCLoss",
  "",
  "surr_name = sys.argv[1]",
  "csv_path  = sys.argv[2]",
  "",
  "surrogates = {",
  '    "squared":       {"hparam": 1.0},',
  '    "squared_hinge": {"hparam": 1.0},',
  '    "hinge":         {"hparam": 1.0},',
  '    "logistic":      {"hparam": 2.0},',
  '    "barrier_hinge": {"hparam": (1.0, 2.0)},',
  "}",
  "",
  "with open(csv_path) as f:",
  "    reader = csv.reader(f)",
  "    rows = list(reader)",
  "",
  "y_pred = torch.tensor([float(x) for x in rows[0]], dtype=torch.float32)",
  "y_true = torch.tensor([float(x) for x in rows[1]], dtype=torch.float32)",
  "",
  "cfg = surrogates[surr_name]",
  "loss_fn = PairwiseAUCLoss(surr_loss=surr_name, hparam=cfg['hparam'])",
  "result = loss_fn(y_pred.view(-1, 1), y_true.view(-1, 1)).item()",
  "print(f'{result:.15g}')"
))
```

## R wrapper that calls the Python subprocess

```{r}
call_py_loss <- function(surr_name, y_pred_r, y_true_r) {
  csv_path <- tempfile(fileext = ".csv")
  writeLines(c(
    paste(sprintf("%.15g", as.numeric(y_pred_r)), collapse = ","),
    paste(as.integer(y_true_r), collapse = ",")
  ), csv_path)
  
  # Call the isolated Python binary from reticulate environment
  out <- system2(py_bin, c(py_loss_script, surr_name, csv_path),
                 stdout = TRUE, stderr = TRUE)
                 
  # Ensure we capture Python errors properly if they happen
  res <- suppressWarnings(as.numeric(out[length(out)]))
  if (is.na(res)) stop("Python script failed: ", paste(out, collapse="\n"))
  res
}
```

## Sample Input Data

Let's generate some sample predictions and labels to use for a quick demonstration:

```{r}
torch::torch_manual_seed(42L)
N <- 10L
n_pos <- 3L
n_neg <- 7L

y_pred_r <- torch::torch_randn(N, dtype = torch::torch_float32())
y_true_r <- torch::torch_tensor(
  c(rep(1, n_pos), rep(0, n_neg)), dtype = torch::torch_float32()
)

cat("Predictions:\n")
print(y_pred_r)
cat("Labels:\n")
print(y_true_r)
```

## R implementation

Below is the R implementation of PairwiseAUCLoss from python libauc.

```{r}
surrogate_loss_names <- c(
  "squared",
  "squared_hinge",
  "hinge",
  "logistic",
  "barrier_hinge"
)

squared_loss <- function(margin, t) {
  (margin - t)^2
}

squared_hinge_loss <- function(margin, t) {
  torch::torch_clamp(margin - t, min = 0)^2
}

hinge_loss <- function(margin, t) {
  torch::torch_clamp(margin - t, min = 0)
}

logistic_loss <- function(scale, t) {
  torch::nnf_softplus(-scale * t)
}

barrier_hinge_loss <- function(margin, scale, t) {
  torch::torch_maximum(
    -scale * (margin + t) + margin,
    torch::torch_maximum(margin - t, scale * (t - margin))
  )
}

get_surrogate_loss <- function(loss_name = "squared_hinge") {
  switch(loss_name,
    squared = squared_loss,
    squared_hinge = squared_hinge_loss,
    hinge = hinge_loss,
    logistic = logistic_loss,
    barrier_hinge = barrier_hinge_loss,
    stop(sprintf(
      "Unknown surrogate loss '%s'. Choose from: %s",
      loss_name,
      paste(surrogate_loss_names, collapse = ", ")
    ), call. = FALSE)
  )
}

nn_pairwise_auc_loss <- torch::nn_module(
  c("PairwiseAUC","nn_loss"),
  initialize = function(
      surr_loss = "logistic",
      margin = 1.0,
      scale = 1.0) {
    self$surr_loss_name <- surr_loss
    self$margin <- margin
    self$scale <- scale
    self$surrogate <- get_surrogate_loss(surr_loss)
    for (name in c("evals", "zeros", "all_one_class")) {
      self$buffer(name)
    }
  },
  buffer = function(name, value = torch::torch_tensor(0L)) {
    self[[name]] <- torch::nn_buffer(value)
  },
  increment = function(name) {
    self$buffer(name, self[[name]] + 1L)
  },
  forward = function(input, target) {
    self$increment("evals")
    if (torch::as_array((target[1] == target)$all())) {
      self$increment("all_one_class")
      self$increment("zeros")
      return(torch::torch_sum(input * 0))
    }
    y_pred <- input$flatten()
    y_true <- target$flatten()
    pos_mask <- y_true == 1L
    neg_mask <- y_true == 0L
    f_ps <- y_pred[pos_mask]
    f_ns <- y_pred[neg_mask]
    t <- f_ps$unsqueeze(2) - f_ns$unsqueeze(1)
    if (self$surr_loss_name == "barrier_hinge") {
      loss <- self$surrogate(self$margin, self$scale, t)
    } else if (self$surr_loss_name == "logistic") {
      loss <- self$surrogate(self$scale, t)
    } else {
      loss <- self$surrogate(self$margin, t)
    }
    loss_val <- loss$mean()
    if (torch::as_array(loss_val == 0)) {
      self$increment("zeros")
    }
    loss_val
  }
)
```

Below is our R wrapper that calls `nn_pairwise_auc_loss` with the same
hyperparameters used in the Python code above.

```{r}
r_loss <- function(surr_name, y_pred, y_true) {
  hparams <- list(
    squared       = list(margin = 1.0),
    squared_hinge = list(margin = 1.0),
    hinge         = list(margin = 1.0),
    logistic      = list(scale = 2.0),
    barrier_hinge = list(margin = 1.0, scale = 2.0)
  )
  args <- c(list(surr_loss = surr_name), hparams[[surr_name]])
  loss_fn <- do.call(nn_pairwise_auc_loss, args)
  as.numeric(loss_fn(y_pred, y_true)$item())
}
```

## Verification with atime

We use `atime` to compute both Python and R loss values across
increasing data sizes, with `result=TRUE` so we can compare the actual
outputs using `all.equal`.

```{r}
torch::torch_manual_seed(42L)
surr_names <- c("squared", "squared_hinge", "hinge", "logistic", "barrier_hinge")
n_seq <- as.integer(10^seq(1, 3, by = 0.5))
result_list <- list()

for (surr in surr_names) {
  bench <- atime::atime(
    N = n_seq,
    result = TRUE,
    seconds.limit = 10,
    setup = {
      torch::torch_manual_seed(N)
      n_pos <- max(1L, as.integer(N * 0.3))
      n_neg <- N - n_pos
      y_pred_r <- torch::torch_randn(N, dtype = torch::torch_float32())
      y_true_r <- torch::torch_tensor(
        c(rep(1, n_pos), rep(0, n_neg)), dtype = torch::torch_float32()
      )
    },
    R = {
      r_loss(surr, y_pred_r, y_true_r)
    },
    Python = {
      call_py_loss(surr, y_pred_r, y_true_r)
    }
  )
  compare_dt <- data.table::dcast(
    bench$measurements, N ~ expr.name, value.var = "result"
  )[
    , equal := paste(all.equal(as.numeric(R[[1]]), as.numeric(Python[[1]])))
    , by = N
  ]
  result_list[[surr]] <- compare_dt[, .(surrogate = surr, N, equal)]
}
all_results <- data.table::rbindlist(result_list)
print(all_results)
```

The table above shows `TRUE` for each (surrogate, N) pair where R and
Python return the same value, or a small relative difference for large N
(normal float32 rounding). This confirms our R implementation matches
the original Python libauc.

## Source code

The full R implementation will be available in
[pairwise_auc_loss.R](https://github.com/mlr3-imbalanced/mlr3torchAUM/blob/pairwise-auc-loss/R/pairwise_auc_loss.R).
The Python reference is
[libauc PairwiseAUCLoss](https://github.com/Optimization-AI/LibAUC/blob/main/libauc/losses/auc.py#L855)
with
[surrogate losses](https://github.com/Optimization-AI/LibAUC/blob/main/libauc/losses/surrogate.py).
