---
title: "Softmax, Cross Entropy, and Hellinger Loss: A Numerical-Stability Tour"
author: "Wei Cao"
date: "2026-06-25"
categories: [R, torch, numerical-stability]
description: "Forward and backward numerical stability for softmax-based losses."
execute:
  echo: true
  warning: true
  error: true
---

## Goal

The goal of this post is to discuss numerical-stability issues in the forward and backward propagation of several loss functions that involve softmax, and to try to propose some solutions.
In particular, the losses discussed here are Cross Entropy and a Hellinger-distance loss.

Anti-goal:
the architecture of the classifier neural network itself. That is outside the scope of this post.

Goal for future posts:
for the numerical-stability solutions proposed in this post, run empirical analyses on multiple datasets and at multiple scales. It is possible that more elegant new solutions will be born in that process.
Following the final question raised at the end of this post, future posts will also explore the training dynamics of loss functions. That analysis will also be combined with empirical results. Stay tuned.

## 1. Overflow and Underflow: What We Need to Know About Float32

Float32 is the data type we commonly use for training machine-learning models, and it is also the default data type for torch tensors.
From its construction, we can answer the following questions. I omit the calculation process here; if needed, I can write a separate post about it in the future.

1. What is the largest representable number? \(2^{128} - 2^{104}\), approximately \(3.40 \times 10^{38}\). Slightly larger than this, and so on below.
2. What is the smallest positive normal number? \(2^{-126}\), approximately \(1.17 \times 10^{-38}\).
3. What is the smallest positive subnormal number? \(2^{-149}\), approximately \(1.40 \times 10^{-45}\).
4. What is the spacing between two neighboring float32 numbers near 1? \(2^{-23}\), approximately \(1.19 \times 10^{-7}\).
5. What is the spacing between neighboring float32 numbers around any given number? \(2^{e-23}\). For example, near \(2^{23}\) (8,388,608), the spacing between neighboring numbers is 1.

```{r}
library(torch)
big <- torch_tensor(3.40e38, dtype=torch_float32())
bigger <- torch_tensor(3.41e38, dtype=torch_float32())
cat("big enough: ", as.numeric(big), "\n") # 3.4e38
cat("bigger than big: ", as.numeric(bigger), "\n") # Inf

small <- torch_tensor(1.41e-45, dtype=torch_float32())
smaller <- torch_tensor(1.40e-45, dtype=torch_float32()) # surprisingly displayed the same as small above
really_small <- torch_tensor(0.70e-45, dtype=torch_float32()) # done for; rounded to zero
cat("small enough: ", as.numeric(small), "\n")
cat("smaller than small: ", as.numeric(smaller), "\n")
cat("really small: ", as.numeric(really_small), "\n")

one_delta <- torch_tensor(1 + 1.19e-7, dtype=torch_float32())
one_delta_plus <- torch_tensor(1 + 6e-7, dtype=torch_float32())
cat("one + delta: ", as.numeric(one_delta), "\n") # still 1
cat("one + delta plus: ", as.numeric(one_delta_plus), "\n") # 1.000001 = 1 + e-6

number <- torch_tensor(2^23, dtype=torch_float32())
number_delta05 <- torch_tensor(2^23+0.5, dtype=torch_float32())
number_delta1 <- torch_tensor(2^23+1, dtype=torch_float32())
cat("number: ", as.numeric(number), "\n") # 8388608
cat("number + delta 0.5: ", as.numeric(number_delta05), "\n") # 8388608
cat("number + delta 1: ", as.numeric(number_delta1), "\n") # 8388609
```
All of this is only to give the reader an intuition: at some point, float32 gives up on very large numbers or numbers that get infinitely close to 0, surrendering them as `Inf` or `0`.
Also, the larger a number is, the less precise float32 becomes, and the spacing between neighboring float32 values becomes larger.

## 2. The Truth About Cross Entropy and Softmax

Now our setting is as follows: we face a multiclass-classification task with \(K\) classes. The classifier outputs logits, and we use softmax to convert them into probabilities.
If \(K = 3\), then \((z_1, z_2, z_3) \rightarrow (p_1, p_2, p_3)\), where \(p_1 + p_2 + p_3 = 1\).
If class 1 is the true class, then the one-hot vector representing the ground truth is \((1, 0, 0)\).

We call \(p_1\) \(p_t\), and temporarily refer to both \(p_2\) and \(p_3\) as \(p_f\).
We call \(z_1\) \(z_t\), and temporarily refer to both \(z_2\) and \(z_3\) as \(z_f\).
We use \(p_k\) to refer collectively to \(p_1, p_2, p_3\). You may also see other subscripts, such as \(p_i\); they are only names and have no special meaning.
Similarly, we use \(z_k\) to refer collectively to \(z_1, z_2, z_3\). You may also see other subscripts, such as \(z_i\); they are only names and have no special meaning.

### 2.1 Cross Entropy Formula

It is simply \(-\log(p_t)\).
I will not discuss its statistical or information-theoretic interpretations here; for now, it is enough to know the formula.

### 2.2 Prerequisites

#### 2.2.1 The Softmax Formula

$$
p_k = \frac{\exp(z_k)}{\sum_{i=1}^{K}\exp(z_i)}.
$$

#### 2.2.2 The Jacobian of Softmax

This is the partial derivative of the probability \(p_k\), after softmax conversion, with respect to logit \(z_i\). Each pair \((k, i)\) corresponds to one entry in the Jacobian matrix.

$$
J_{ki} = \frac{\partial p_k}{\partial z_i}.
$$

Again using the three-class task (\(K=3\)) as an example, the Jacobian is:

$$
\frac{\partial p_1}{\partial z_1}, \frac{\partial p_1}{\partial z_2}, \frac{\partial p_1}{\partial z_3};
$$

$$
\frac{\partial p_2}{\partial z_1}, \frac{\partial p_2}{\partial z_2}, \frac{\partial p_2}{\partial z_3};
$$

$$
\frac{\partial p_3}{\partial z_1}, \frac{\partial p_3}{\partial z_2}, \frac{\partial p_3}{\partial z_3}.
$$

Now that we have the softmax formula, we can compute this Jacobian.

$$
\frac{\partial p_k}{\partial z_i}
= \frac{\exp(z_k)'\sum_{i=1}^{K}\exp(z_i) - \left(\sum_{i=1}^{K}\exp(z_i)\right)'\exp(z_k)}
{\left(\sum_{i=1}^{K}\exp(z_i)\right)^2}.
$$

When \(k=i\), \(\exp(z_k)' = \exp(z_k)\), and \(\sum_{i=1}^{K}\exp(z_i)' = \exp(z_k)\).
Therefore:

$$
\frac{\partial p_k}{\partial z_i}
= \frac{\exp(z_k)\sum_{i=1}^{K}\exp(z_i) - \exp(z_k)\exp(z_k)}
{\left(\sum_{i=1}^{K}\exp(z_i)\right)^2}
= p_k - p_k^2
= p_k(1-p_k).
$$

When \(k \neq i\), \(\exp(z_k)' = 0\), and \(\sum_{i=1}^{K}\exp(z_i)' = \exp(z_i)\).
Therefore:

$$
\frac{\partial p_k}{\partial z_i}
= \frac{0 \cdot \sum_{i=1}^{K}\exp(z_i) - \exp(z_i)\exp(z_k)}
{\left(\sum_{i=1}^{K}\exp(z_i)\right)^2}
= -p_k p_i.
$$

For a three-class task, the softmax Jacobian is:

$$
\begin{pmatrix}
p_1(1-p_1) & -p_1p_2 & -p_1p_3 \\
-p_2p_1 & p_2(1-p_2) & -p_2p_3 \\
-p_3p_1 & -p_3p_2 & p_3(1-p_3)
\end{pmatrix}.
$$

Now abstract it one level further with the Kronecker delta:

$$
\delta_{ij} =
\begin{cases}
1, & \text{if } i = j \\
0, & \text{if } i \neq j
\end{cases}
$$

Then:

$$
J_{ki} = p_k(\delta_{ki}-p_i).
$$

### 2.3 The Gradient of Cross Entropy

We use \(L\) to denote Cross Entropy, or any other loss function.
By the chain rule:

$$
\frac{\partial L}{\partial z_i}
= \sum_{k=1}^{K}
\left(
\frac{\partial L}{\partial p_k}
\frac{\partial p_k}{\partial z_i}
\right).
$$

Since the Cross Entropy formula \(-\log(p_t)\) only depends on \(p_t\):

$$
\frac{\partial L}{\partial z_i}
= \frac{\partial L}{\partial p_t}
\frac{\partial p_t}{\partial z_i}.
$$

\(\frac{\partial L}{\partial p_t}\) is easy to compute: it is \(-1/p_t\).
\(\frac{\partial p_t}{\partial z_i}\) can be read from the softmax Jacobian.

$$
\frac{\partial L}{\partial z_i}
= -\frac{1}{p_t}J_{ti}
= -\frac{1}{p_t}p_t(\delta_{ti}-p_i)
= p_i-\delta_{ti}.
$$

For the true class, \(t=i\):

$$
\frac{\partial L}{\partial z_t} = p_t - 1.
$$

For a non-true class, \(t \neq i\):

$$
\frac{\partial L}{\partial z_f} = p_f.
$$

Besides the representation \(p_i-\delta_{ti}\), which uses the Kronecker delta, a second representation is vector-based:
the softmax gradient is

$$
p-y,
$$

where \(y\) is the ground-truth one-hot vector.

For example, in the three-class case where class 1 is true, the gradient is:

$$
(p_1,p_2,p_3)-(1,0,0)=(p_1-1,p_2,p_3).
$$

These are then passed as upstream gradients to each parameter of the classifier neural network.

This is quite natural. For the true class, the gradient is \(p_1-1\), always negative; the smaller \(p_1\) is, the more negative it is, pulling \(p_1\) upward until it approaches 1.
For a non-true class, the gradient is always positive, pulling \(p_f\) downward until it approaches 0.

### 2.4 Computing Cross Entropy and Its Gradient Naively

The naive way to compute Cross Entropy is to apply `-log` on top of the true-class softmax probability.

$$
\mathrm{CE}=-\log(p_t)
=-\log\left(\frac{\exp(z_t)}{\sum_i \exp(z_i)}\right).
$$

This is the forward computation. It can actually be transformed with algebraic identities; I will introduce that later.

For backward propagation, the gradient is what we introduced earlier:

$$
\frac{\partial L}{\partial z_i}
= \frac{\partial L}{\partial p_t}
\frac{\partial p_t}{\partial z_i}
= -\frac{1}{p_t}p_t(\delta_{ti}-p_i).
$$

Why did I not simplify the expression above?

**Because autograd does not simplify it for us!**

When we compute Cross Entropy naively, the inner softmax is one operator, with its own forward and backward computation rules; the outer `-log` is another independent operator, with its own forward and backward computation rules.
The outer and inner operators are mechanically connected by the chain rule. They do not need to know about each other's existence, like two upstream and downstream manufacturers in a global supply chain: they may be in different countries and do not interfere with each other's company operations.

In the expression \(1/p_t \cdot p_t \cdot (\delta_{ti}-p_i)\), \(-1/p_t\) is the gradient of the outer merchant, `-log`; \(p_t(\delta_{ti}-p_i)\) is the gradient of the inner merchant, softmax.
They do not interfere with each other's company operations, so they do not cancel.

### 2.5 A Numerical-Stability Checkup for Naive Cross Entropy

After introducing so much about Cross Entropy and softmax, it may seem that we forgot why we studied those important facts about float32 and ran all those experiments earlier.
It was because we need to study numerical stability.
So let us start the analysis now.
When analyzing it, we must not miss either forward propagation or backward propagation.

As mentioned earlier, the naive way to compute Cross Entropy is to apply `-log` on top of the true-class softmax probability.
Therefore, our analysis matrix is a two-by-two combination: softmax plus `-log`, forward plus backward, for four dimensions in total:

1. Forward softmax.
2. Forward `-log`.
3. Backward softmax.
4. Backward `-log`.

#### 2.5.1 Forward Softmax

Because softmax involves the exponential function, we always worry that a very large logit may produce overflow.
Indeed, this is a very natural concern.

```{r}
zmax <- torch_tensor(89, dtype = torch_float32())
cat("torch exp(100): ", as.numeric(torch_exp(zmax)), "\n")
```
Empirically, for float32, when a logit reaches 89, applying the natural exponential turns it into `Inf`.
But for softmax, we can do a simple algebraic identity transformation.

$$
p_k
= \mathrm{softmax}(z_k)
= \frac{\exp(z_k)}{\sum_{i=1}^{K}\exp(z_i)}
= \frac{\exp(z_k)\exp(-z_{\max})}{\sum_{i=1}^{K}\exp(z_i)\exp(-z_{\max})}.
$$

That is, multiply both numerator and denominator by \(\exp(-z_{\max})\).
Then:

$$
p_k
= \frac{\exp(z_k-z_{\max})}{\sum_{i=1}^{K}\exp(z_i-z_{\max})}.
$$

Therefore, the forward computation of softmax cannot overflow, as long as the logits themselves do not overflow.

```{r}
big <- torch_tensor(3.40e38, dtype=torch_float32(), requires_grad = TRUE)
small <- torch_tensor(-3.40e38, dtype=torch_float32(), requires_grad = TRUE)
zero <- torch_tensor(0, dtype = torch_float32(), requires_grad = TRUE)

cat("big after subtracting the max logit: ", as.numeric(big - big), "\n")
cat("small after subtracting the max logit: ", as.numeric(small - big), "\n")
cat("zero after subtracting the max logit: ", as.numeric(zero - big), "\n")
exp_big_subtracted <- torch_exp(big - big)
exp_small_subtracted <- torch_exp(small - big)
exp_zero_subtracted <- torch_exp(zero - big)
sum_exp <- exp_big_subtracted + exp_small_subtracted + exp_zero_subtracted
cat("big after subtracting the max logit and exponentiating: ", as.numeric(exp_big_subtracted), "\n")
cat("small after subtracting the max logit and exponentiating: ", as.numeric(exp_small_subtracted), "\n")
cat("zero after subtracting the max logit and exponentiating: ", as.numeric(exp_zero_subtracted), "\n")
cat("denominator", as.numeric(sum_exp <- exp_big_subtracted + exp_small_subtracted + exp_zero_subtracted), "\n")

p_big <- exp_big_subtracted/sum_exp
p_small <- exp_small_subtracted/sum_exp
p_zero <- exp_zero_subtracted/sum_exp
cat("p_big: ", as.numeric(p_big), "\n")
cat("p_small: ", as.numeric(p_small), "\n")
cat("p_zero: ", as.numeric(p_zero), "\n")
```
We can see that no overflow occurred.
The denominator of softmax is at least 1, so forward softmax computation is very safe.

Although no overflow occurred, underflow did occur: after subtracting this `big`, other logits may become extremely small and directly underflow to negative infinity.
They may not even need to be that small; after exponentiation, they may underflow to 0.
This plants a hidden danger for the backward propagation of `log`.

#### 2.5.2 Backward Softmax

To compute backward softmax, we only need to substitute into the softmax Jacobian:

$$
J_{ki}=p_k(\delta_{ki}-p_i).
$$

Here we assume class 2 is the true class: confidently wrong.
Because Cross Entropy only uses the true-class \(p_t\), we only examine this \(p_t\), which is `p_small`.

$$
J_{21}=-p_2p_1=-0\cdot 1=0,
$$

$$
J_{22}=p_2(1-p_2)=0\cdot(1-0)=0,
$$

$$
J_{23}=-p_2p_3=-0\cdot 0=0.
$$

```{r}
p_small$backward()
cat("gradient passed from p_small, the true-class pt, to the big logit: ", as.numeric(big$grad), "\n")
cat("gradient passed from p_small, the true-class pt, to the small logit: ", as.numeric(small$grad), "\n")
cat("gradient passed from p_small, the true-class pt, to the zero logit: ", as.numeric(zero$grad), "\n")
```
#### 2.5.3 Forward `-log`

The risk of non-finite values in the forward computation of `log` comes from 0 and negative numbers.
However, \(p_k\) cannot be negative; underflowing to 0 is very possible.
Since Cross Entropy only takes the true-class softmax probability \(p_t\), i.e. \(-\log(p_t)\), when does it become 0?
When the model is confidently wrong.

```{r}
pt_conf_wrong <- torch_tensor(0, dtype = torch_float32())
ce <- -log(pt_conf_wrong)
cat("Cross Entropy when confidently wrong: ", as.numeric(ce), "\n")
```
This is not actually that harmful; it is expected.
But what happens to backward propagation in this situation?

#### 2.5.4 Backward `-log`

We know that the derivative of \(\log(x)\) is \(1/x\).
With the negative sign in Cross Entropy, it becomes \(-1/p_t\).

By the chain rule:

For the gradient passed from CE to \(z_t\), the true-class logit:

$$
-\frac{1}{p_t}p_t(1-p_t).
$$

For the gradient passed from CE to \(z_f\), a non-true-class logit:

$$
-\frac{1}{p_t}p_t(-p_f).
$$

Although we have a strong urge to simplify and cancel terms, we already said that because we are computing Cross Entropy naively, the `log` operator and the softmax operator are independent operators, and gradient computation is fully handed over to autograd.
We emphasized earlier that **autograd does not automatically cancel terms**. It is hard to overemphasize this.

This \(p_t\) is the root cause of NaN in backward propagation.
When \(p_t\) approaches 0, meaning the model is confidently wrong, \(-1/p_t\) becomes \(-\mathrm{Inf}\). The later multiplier \(p_t\) is also 0.
An `Inf` multiplied by 0 naturally becomes `NaN`.

```{r}
z <- torch_tensor(c(-1000, 1000, 1000), dtype = torch_float32(), requires_grad = TRUE) # true class is 1; confidently wrong
p <- nnf_softmax(z, dim=1)
l <- - torch_log(p[1])
l$backward()
z$grad
```
The result confirms our analysis.
The entire army is wiped out; none is spared.

### 2.6 Solution: Invent a New Operator

Now we can finally stop suppressing our urge to simplify.

In fact, we already simplified it earlier in Section 2.3.
The conclusion:
the gradient passed from Cross Entropy to the logits is:

$$
p_i-\delta_{ti}.
$$

In vector form, it is \(p-y\), where \(y\) is the ground-truth one-hot vector.

This is a very clean result, without any troublesome denominator.

Therefore, we invent a combined `log + softmax` operator and directly use the simplified gradient formula for backward propagation.

Although we have already analyzed it in detail, let us reorganize the reasoning with more standard notation.

When computing the gradient of the softmax layer, we focus on the partial derivative of the loss function \(L\) with respect to input \(z_k\):

$$
\frac{\partial L}{\partial z_k}
= \sum_{i=1}^{N}
\frac{\partial L}{\partial y_i}
\frac{\partial y_i}{\partial z_k}.
$$

Here:

- \(\mathrm{grad}_i\) represents \(\frac{\partial L}{\partial y_i}\).
- \(\delta_{ik}\) is the Kronecker delta: 1 when \(i=k\), otherwise 0.
- \(p_k\) is the softmax output probability.

To understand the derivation below, we must pay attention to what \(y\) refers to here.

In the previous analysis, we split `-log` and softmax into two independent functions: softmax inside and `-log` outside. Together they form Cross Entropy.
So our setting was always: \(y\) is \(-\log(x)\), and \(x\) is the result of softmax.
Thus our chain rule was always: derivative of the outer `-log` multiplied by the Jacobian of the inner softmax.

But now \(y\) has changed, because we combine `log` and softmax into one operator, `log_softmax`.
So Cross Entropy is simply `-log_softmax`; \(y\) is very simple: \(-x\), where \(x\) is the result of `log_softmax`.

Since the gradient of Cross Entropy is \(p_i-\delta_{ti}\), adding a negative sign gives the gradient of `log_softmax`:

$$
\delta_{ti}-p_i.
$$

\(\frac{\partial L}{\partial y_i}\) is actually extremely simple: it is \(-1\) for the true-class `log_softmax`, and 0 otherwise.
But we will not expand that here.

Now let us do some algebra.

Expand the computation:

$$
\frac{\partial L}{\partial z_k}
= \sum_{i=1}^{N}\mathrm{grad}_i(\delta_{ik}-p_k).
$$

Use distributivity to split the summation:

$$
\sum_i \mathrm{grad}_i(\delta_{ik}-p_k)
= \sum_i(\mathrm{grad}_i\delta_{ik})
- \sum_i(\mathrm{grad}_i p_k).
$$

Final simplified form:

Since \(\delta_{ik}\) is 0 when \(i \neq k\), the summation \(\sum_i(\mathrm{grad}_i\delta_{ik})\) leaves only the \(i=k\) term, namely \(\mathrm{grad}_k\).
At the same time, pulling out the constant \(p_k\) gives:

$$
\frac{\partial L}{\partial z_k}
= \mathrm{grad}_k
- p_k\sum_{i=1}^{N}\mathrm{grad}_i.
$$

This formula expresses the concise form of **softmax backward propagation**:
the gradient of the current node equals its upstream gradient, minus an offset term produced by weighting and summing all gradients.

Remember this formula. It will help us read the `log_softmax` source code.

### 2.7 A Journey Through the `log_softmax` Source Code

The exciting moment has finally arrived.
How does autograd actually know how to compute these gradients during backward propagation?
We memorize derivative formulas for each function and calculate them by hand.

For autograd, we give it a cheat sheet:
<https://github.com/pytorch/pytorch/blob/v2.2.2/tools/autograd/derivatives.yaml#L2089>

```yaml
- name: _log_softmax(Tensor self, int dim, bool half_to_float) -> Tensor
  self: _log_softmax_backward_data(grad, result, dim, self.scalar_type())
  result: self_t - logsumexp_jvp(self_p, self_t, {dim}, true)
```

Compare this with the life-and-death book for some other operators:

```yaml
- name: log(Tensor self) -> Tensor
  self: grad.div(self.conj())
```

`grad` is the upstream gradient. According to the chain rule, it appears here as a multiplier.
`self` is the input data, the \(x\) in \(\log(x)\).
`conj()` exists for the complex-number case; we can ignore it for now.
`div` is division.

This matches the derivative formula for \(\log(x)\), namely \(1/x\).

Compare it with `sqrt`, the protagonist of the next section:

```yaml
- name: sqrt(Tensor self) -> Tensor
  self: grad / (2 * result.conj())
  result: auto_element_wise
```

`result` is the forward output of this operator, i.e. \(\sqrt{x}\) rather than \(x\).
This also matches the derivative formula for \(\sqrt{x}\): \(1/(2\sqrt{x})\).

But here, the backward propagation of `log_softmax` is delegated to a function called `_log_softmax_backward_data`.

Finding its true body takes some effort, because PyTorch needs to consider different architectures and instruction sets, among other engineering concerns.
Following the clue `_log_softmax_backward_data`, we find this in:
<https://github.com/pytorch/pytorch/blob/main/aten/src/ATen/native/native_functions.yaml#L3733>

```yaml
- func: _log_softmax_backward_data.out(Tensor grad_output, Tensor output, int dim, ScalarType input_dtype, *, Tensor(a!) out) -> Tensor(a!)
  structured: True
  dispatch:
    CPU: log_softmax_backward_cpu_out
    CUDA: log_softmax_backward_cuda_out
    MTIA: log_softmax_backward_mtia_out
    MPS: log_softmax_backward_mps_out
    XPU: log_softmax_backward_xpu_out
```

Using CPU as the example, we get `log_softmax_backward_cpu_out`.
Where do we find it?
In <https://github.com/pytorch/pytorch/blob/main/aten/src/ATen/native/SoftMax.cpp#L386>,
inside `TORCH_IMPL_FUNC(log_softmax_backward_cpu_out)`, we have:

```cpp
if (grad_.ndimension() > 0 && dim_ == grad_.ndimension() - 1) {
  log_softmax_backward_lastdim_kernel(kCPU, grad_input, grad_, output_);
} else {
  log_softmax_backward_kernel(kCPU, grad_input, grad_, output_, dim_);
}
```

Our softmax is on the final dimension of the tensor, so we enjoy some special acceleration design and take the `log_softmax_backward_lastdim_kernel` path.

Then, in <https://github.com/pytorch/pytorch/blob/main/aten/src/ATen/native/cpu/SoftMaxKernel.cpp#L1064>, we find:

```cpp
ALSO_REGISTER_AVX512_DISPATCH(
    log_softmax_backward_lastdim_kernel,
    &log_softmax_backward_lastdim_kernel_impl)
```

This registers it to the CPU AVX512 instruction set. In principle, there should also be a fallback DISPATCH registration, but I did not find it at the moment.
These details do not affect our source-code reading.

Like treasure hunters, we continue along this clue:
<https://github.com/pytorch/pytorch/blob/main/aten/src/ATen/native/cpu/SoftMaxKernel.cpp#L1013>

```cpp
void log_softmax_backward_lastdim_kernel_impl(
    const Tensor& grad_input,
    const Tensor& grad,
    const Tensor& output) {
  AT_DISPATCH_FLOATING_TYPES_AND2(
      at::ScalarType::BFloat16, at::ScalarType::Half, grad.scalar_type(),
      "log_softmax_backward_lastdim_kernel_impl", [&] {
        vec_host_softmax_backward_lastdim<scalar_t, true>::apply(
            grad_input, grad, output);
      });
}
```

This DISPATCH adapts Python's dynamic types to C++ static compilation and supports multiple floating-point precisions.
Eventually it passes control to `vec_host_softmax_backward_lastdim`.
`scalar_t` is the data type; in our case, for example, it is float32.
The `true` argument exists because softmax and `log_softmax` use different branches of the same function, so `true` tells it that we are computing `log_softmax` rather than softmax.

This function is at:
<https://github.com/pytorch/pytorch/blob/main/aten/src/ATen/native/cpu/SoftMaxKernel.cpp#L913>

Finally, at:
<https://github.com/pytorch/pytorch/blob/main/aten/src/ATen/native/cpu/SoftMaxKernel.cpp#L162>

after many twists and turns, we finally find the source implementation we wanted.

```cpp
if constexpr (log_softmax) {
  auto sum = vec::reduce_all<scalar_t>(
      [](Vec& x, Vec& y) { return x + y; }, grad_data, dim_size);
  vec::map2(
      [sum](Vec x, Vec y) { return x - ((y.exp()) * Vec(sum)); },
      grad_input_data,
      grad_data,
      output_data,
      dim_size);
} else {
  // Softmax
}
```

The `else` branch is the softmax branch. We look at `log_softmax`.
Here:

- `grad_data` \(=\) the upstream gradient \((\mathrm{d}L/\mathrm{d}\text{ output})\).
- `output_data` \(=\) the forward output: \(\log p\) in the `log_softmax` branch, and \(p\) in the softmax branch.
- `grad_input_data` \(=\) the result to be written in \((\mathrm{d}L/\mathrm{d}\text{ logits } z)\).
- `dim_size` \(=\) the number of classes \(K\); `scalar_t` \(=\) the element type, such as float or double.

```cpp
auto sum = vec::reduce_all<scalar_t>(
    [](Vec& x, Vec& y) { return x + y; }, grad_data, dim_size);
```

This line first takes `grad_data`, the upstream gradient, and sums it:
\(\sum_i^K \mathrm{grad}_i\).

Next:

```cpp
vec::map2(
    [sum](Vec x, Vec y) { return x - ((y.exp()) * Vec(sum)); },
    grad_input_data, // output
    grad_data,       // x
    output_data,     // y
    dim_size);
```

`[sum]` captures the `sum` computed above, i.e. \(\sum_i^K \mathrm{grad}_i\).
`[sum](Vec x, Vec y) { return x - ((y.exp()) * Vec(sum)); }` is an anonymous function, meaning that this operation is applied elementwise to vectors \(x\) and \(y\).

`grad_data` is the upstream gradient \(\mathrm{grad}_i\), and it is the anonymous function's `Vec x`.
`output_data` is \(\log(p)\), which in the three-class case can be understood as the vector \((\log(p_1), \log(p_2), \log(p_3))\).
`y.exp()` applies `exp` to \(\log(p_i)\), turning it back into \(p_i\).
`Vec(sum)` broadcasts the upstream-computed sum into a vector and multiplies it elementwise by `y.exp()`, i.e. \(p_i\).

Putting it all together, each element of the computed gradient vector is:

$$
\mathrm{grad}_i - p_i \sum_i^K \mathrm{grad}_i.
$$

If you flip back to Section 2.6, "Solution: Invent a New Operator", this is word for word the same.

### 2.8 Postscript

Incidentally, in some textbooks we often see another algebraic identity for Cross Entropy:

$$
\mathrm{CE}
=-\log(p_t)
=-\log\left(\frac{\exp(z_t)}{\sum_i \exp(z_i)}\right)
=-z_t+\mathrm{logsumexp}(z_i).
$$

If we take its derivative with respect to \(z_k\), what do we get?
\(-z_t\) gives \(-\delta_{tk}\).
The derivative of \(\mathrm{logsumexp}(z_i)\) with respect to \(z_k\) is our old friend \(\mathrm{softmax}(z_i)\), i.e. \(p_i\).
Together, this gives \(p_i-\delta_{tk}\).
Different paths, same destination.

## 3. The Truth About Hellinger

Now, after reading a certain paper, we learn that it uses a metric based on Hellinger Distance as a multiclass evaluation metric.
We want to build a new loss function based on Hellinger Distance.
We also face numerical-stability issues.

The formula for Hellinger Distance is:

$$
\frac{\sqrt{\sum_{i=1}^{K}(\delta_{ti}-\sqrt{p_i})^2}}{\sqrt{2}}.
$$

\(K\) is still the total number of classes. \(i\) is only a nominal subscript; later we will also use \(k\), and so on.
The subscript \(t\) still represents the true class.
\(\delta_{ti}\) is still the Kronecker delta: it equals 1 when \(i=t\), and 0 when \(i\neq t\).

That is:

$$
\frac{\lVert \text{one-hot-vector}-\sqrt{p}\rVert_2}{\sqrt{2}},
$$

where \(p\) is the softmax vector.

If we simplify it algebraically:

$$
\frac{\sqrt{\sum_i^K(\delta_{ti}-\sqrt{p_i})^2}}{\sqrt{2}}
= \frac{\sqrt{(1-\sqrt{p_t})^2+\sum_{i\neq t}(\sqrt{p_i})^2}}{\sqrt{2}}.
$$

Then:

$$
= \frac{\sqrt{1-2\sqrt{p_t}+p_t+\sum_{i\neq t}p_i}}{\sqrt{2}}.
$$

Since:

$$
p_t+\sum_{i\neq t}p_i=1,
$$

the expression becomes:

$$
\frac{\sqrt{2-2\sqrt{p_t}}}{\sqrt{2}}
= \sqrt{1-\sqrt{p_t}}.
$$

Very concise.
\(p_t\) is the true-class softmax probability.

### 3.1 The Problem With `sqrt`

Whether we look at the unsimplified Hellinger Distance formula or the simplified Hellinger Distance formula, the moment we see them, we secretly feel something is wrong:

bad news, we have fallen into a nest of square roots.
Backward propagation is dangerous.

As we saw earlier in PyTorch's backward rule for `sqrt`:

```yaml
self: grad / (2 * result.conj())
```

Once \(x \rightarrow 0\), `result`, the forward output \(\sqrt{x}\), also becomes 0, and the whole expression becomes `Inf`.
If the upstream `grad` is 0 at this moment, recall the softmax Jacobian: this is entirely possible.
This is the naked source of `NaN`.

A simple example:

```{r}
s <- torch_tensor(0, dtype = torch_float32(), requires_grad = TRUE)
l <- torch_sqrt(s)
l$backward()
s$grad
```
Now look at the second program:

```{r}
s <- torch_tensor(0, dtype = torch_float32(), requires_grad = TRUE)
l <- torch_sqrt(s*s)
l$backward()
s$grad
```
Let us briefly review why the second program becomes `NaN`.
Let \(y=g(f(x))\), where \(f(x)=x^2\), and \(g(y)=\sqrt{y}\).
Then \(f'(x)=2x\), and \(g'(y)=1/(2\sqrt{y})\).
By the chain rule, the gradient of `s` in the second program is:

$$
\frac{1}{2\sqrt{s^2}} \cdot 2s.
$$

Do you feel the urge to simplify and cancel terms?
For the third time in this post, I emphasize: **autograd does not automatically differentiate algebraically!**

If \(s=0\), or if it is close enough to 0 in the eyes of float32 (recall the smallest positive subnormal number discussed at the beginning), then \(1/(2\sqrt{s^2})\) is `Inf`, and \(s\) is 0.
`Inf * 0` becomes a gorgeous `NaN`.

### 3.2 Re-examining Hellinger Distance

#### 3.2.1 The Original Hellinger Distance Formula

The original formula is a bit tedious to analyze, but it can be done.
The original expression can blow up.
The details are in the analysis below.
Let us walk through the most tedious analysis once; after that, future analyses will not be scary.

Look at:

$$
\sqrt{\sum(\text{one-hot}-\sqrt{p_k})^2}.
$$

What quantities exist under the square root?
We call these the sources of gradient explosion.
Let:

$$
S=\sum(\text{one-hot}-\sqrt{p_k})^2.
$$

This is under the outer square root.
What else appears?
\(p_k\), including two cases: \(p_t\), only one true-class probability, and \(p_f\), multiple non-true-class probabilities, one fewer than the total number of classes.

For the inner \(p_k\), when do we encounter the problem?

- \(p_t \rightarrow 0\) for the true class: confidently wrong.
- \(p_f \rightarrow 0\) for false classes: confidently correct.

For a non-true class:

$$
(0-\sqrt{p_f})^2=(\sqrt{p_f})^2.
$$

Its derivative:

$$
2\sqrt{p_f}\cdot \frac{1}{2\sqrt{p_f}}.
$$

When \(p_f \rightarrow 0\), \(\sqrt{p_f}\) approaches 0, and the derivative of \(\sqrt{p_k}\) approaches infinity.
The result is `NaN`.
When \(p_f \rightarrow 1\), the value is finite.

For the true class:

$$
(1-\sqrt{p_t})^2.
$$

Its derivative:

$$
2(1-\sqrt{p_t})\left(-\frac{1}{2\sqrt{p_t}}\right).
$$

When \(p_t \rightarrow 0\), the expression is \(-\mathrm{Inf}\).
When \(p_t \rightarrow 1\), the expression is 0.

For the outer \(S\):
when our prediction is close to perfect, meaning \((p_1,p_2,p_3)\) is close to the one-hot ground truth, \(S \rightarrow 0\).
Then \(1/(2\sqrt{S})\) is `Inf`; otherwise it is finite.

Putting it together:

Confidently correct:
\(p_t \rightarrow 1\), other \(p_f \rightarrow 0\), the outer layer is `Inf`, the inner non-true classes are already `NaN`, and the inner true-class expression is 0.
`Inf * 0` is `NaN`.
So whichever \(p_k\) we look at, it is `NaN`.

Confidently wrong:
\(p_t \rightarrow 0\), the outer layer is finite, the inner true class has \(-\mathrm{Inf}\), and if a non-true-class \(p_f \rightarrow 0\), the result is `NaN`; if the non-true-class probability is finite or 1, the result is finite.

Verify this with code.

Confidently correct:

```{r}
p <- torch_tensor(c(1, 0, 0), dtype = torch_float32(), requires_grad = TRUE)
oh_confidnet_correct <- torch_tensor(c(1, 0, 0), dtype = torch_float32())
two <- torch_tensor(2, dtype = torch_float32())
hd_confidnet_correct <- torch_sqrt(((oh_confidnet_correct - torch_sqrt(p))^2)$sum())/torch_sqrt(two)
hd_confidnet_correct$backward()
p$grad
```
Perfectly wiped out.
What about when confidently wrong?

```{r}
p <- torch_tensor(c(1, 0, 0), dtype = torch_float32(), requires_grad = TRUE)
oh_confidnet_wrong <- torch_tensor(c(0, 1, 0), dtype = torch_float32()) # class 2 is the true class
two <- torch_tensor(2, dtype = torch_float32())
hd_confidnet_wrong <- torch_sqrt(((oh_confidnet_wrong - torch_sqrt(p))^2)$sum())/torch_sqrt(two)
hd_confidnet_wrong$backward()
p$grad
```
The outer layer is finite, so everything depends on the inner layer.
Our analysis said that the inner true class has \(-\mathrm{Inf}\), and indeed we saw `-Inf`.
For a non-true class with \(p_f \rightarrow 0\), the result is `NaN`, and indeed the third entry is `NaN`.
A non-true class that is finite or equal to 1 has a finite value.

These gradients are then propagated to the logits.
Therefore, whether the model is confidently correct or confidently wrong, both situations produce `NaN`.

#### 3.2.2 The Simplified Hellinger Distance Formula

Now directly minimize:

$$
\sqrt{1-\sqrt{p_t}}.
$$

The derivative is:

$$
-\frac{1}{4\sqrt{p_t}\sqrt{1-\sqrt{p_t}}}.
$$

Thus when \(p_t \rightarrow 0\) or \(p_t \rightarrow 1\), \(-\mathrm{Inf}\) appears.
This formula only considers the true class, so the gradient with respect to \(p_f\) is 0.

Verify with code:

```{r}
p <- torch_tensor(c(1, 0, 0), dtype = torch_float32(), requires_grad = TRUE)
one <- torch_tensor(1, dtype = torch_float32())
hd_confidnet_correct <- torch_sqrt(one-torch_sqrt(p[1])) # class 1 is the true class
hd_confidnet_correct$backward()
p$grad
```
```{r}
p <- torch_tensor(c(1, 0, 0), dtype = torch_float32(), requires_grad = TRUE)
one <- torch_tensor(1, dtype = torch_float32())
hd_confidnet_wrong <- torch_sqrt(one-torch_sqrt(p[2])) # class 2 is the true class
hd_confidnet_wrong$backward()
p$grad
```
Indeed, as expected.

Now include the softmax Jacobian.

For the true class, we need to multiply by \(p_t(1-p_t)\). When \(p_t \rightarrow 0\) or \(p_t \rightarrow 1\), this term is 0.
Together with the previous \(-\mathrm{Inf}\), both confident correctness and confident wrongness produce `NaN`.

For a non-true class, we need to multiply by \(-p_tp_f\).
If \(p_t=0\) or \(p_f=0\), `NaN` will appear.
In other words, both confident correctness and confident wrongness produce `NaN`.
So everything should be wiped out.
Look at the following program.

```{r}
z <- torch_tensor(c(1000, 0, 0), dtype = torch_float32(), requires_grad = TRUE)
p <- nnf_softmax(z, dim=1)
one <- torch_tensor(1, dtype = torch_float32())
hd_confidnet_correct <- torch_sqrt(one-torch_sqrt(p[1])) # class 1 is the true class
hd_confidnet_correct$backward()
z$grad
```
```{r}
z <- torch_tensor(c(1000, 0, 0), dtype = torch_float32(), requires_grad = TRUE)
p <- nnf_softmax(z, dim=1)
one <- torch_tensor(1, dtype = torch_float32())
hd_confidnet_wrong <- torch_sqrt(one-torch_sqrt(p[2])) # class 2 is the true class
hd_confidnet_wrong$backward()
z$grad
```
Correct.

Of course, minimizing \(\sqrt{1-\sqrt{p_t}}\) can be analyzed.
But it is also simply minimizing \(1-\sqrt{p_t}\), i.e. minimizing \(-\sqrt{p_t}\), which makes our life easier.

The derivative is:

$$
-\frac{1}{2\sqrt{p_t}}.
$$

\(p_t\) still inevitably appears in the denominator.
When \(p_t \rightarrow 0\), i.e. when confidently wrong, it becomes \(-\mathrm{Inf}\).

Including the softmax Jacobian:

For the true class:

$$
-\frac{1}{2\sqrt{p_t}}p_t(1-p_t).
$$

For a non-true class:

$$
\frac{1}{2\sqrt{p_t}}p_tp_f.
$$

If we hand this to autograd:

For the true class, when \(p_t \rightarrow 0\):
\(-\mathrm{Inf}\cdot 0 \rightarrow \mathrm{NaN}\).
In the confidently correct case, it is finite.

For a non-true class, when \(p_t \rightarrow 0\), the same logic applies.
In the confidently correct case, it is 0.

In other words, we have reduced the numerically unstable situation to only the case where the model is confidently wrong about the true class.

```{r}
z <- torch_tensor(c(1000, 0, 0), dtype = torch_float32(), requires_grad = TRUE)
p <- nnf_softmax(z, dim=1)
hd_confidnet_correct <- -torch_sqrt(p[1]) # class 1 is the true class
p$retain_grad() # do not release p's grad
hd_confidnet_correct$backward()
cat("p's grad: ", as.numeric(p$grad), "\n")
cat("z's grad: ", as.numeric(z$grad), "\n")
```
When confidently correct, there is no problem.
The gradient of \(z\) becoming 0 is understandable: after all, the prediction is already close to successful.

```{r}
z <- torch_tensor(c(1000, 0, 0), dtype = torch_float32(), requires_grad = TRUE)
p <- nnf_softmax(z, dim=1)
hd_confidnet_correct <- -torch_sqrt(p[2]) # class 2 is the true class
p$retain_grad() # do not release p's grad
hd_confidnet_correct$backward()
cat("p's grad: ", as.numeric(p$grad), "\n")
cat("z's grad: ", as.numeric(z$grad), "\n")
```
This confirms our analysis:
when confidently wrong, it becomes `NaN`.

### 3.3 Solutions

There are several solutions.

#### 3.3.1 Custom Autograd Operator

First, look at the simplest case: minimizing \(-\sqrt{p_t}\).

Return to our calculation.

For the true class:

$$
-\frac{1}{2\sqrt{p_t}}p_t(1-p_t).
$$

For a non-true class:

$$
\frac{1}{2\sqrt{p_t}}p_tp_f.
$$

Please do not suppress the urge to simplify:

$$
-\frac{1}{2\sqrt{p_t}}p_t(1-p_t)
\rightarrow
-\frac{1}{2}(1-p_t)\sqrt{p_t}.
$$

No troublesome denominator.

$$
\frac{1}{2\sqrt{p_t}}p_tp_f
\rightarrow
\frac{1}{2}p_f\sqrt{p_t}.
$$

No troublesome denominator.

Unified, this is:

$$
\frac{\partial l}{\partial z_i}
= \frac{1}{2}(p_i-\delta_{ti})\sqrt{p_t}.
$$

But this introduces another problem: when \(p_t=0\), the gradient is 0, and there is no training force.
Training dynamics are a story for later posts in this series, as mentioned at the beginning of this post.

Because **autograd does not automatically simplify for us** (the fourth time this is mentioned), we need to merge `sqrt` and softmax into one operator.
Earlier, I mentioned that Cross Entropy merges `log` and softmax into one operator; it not only adds a computation rule in the low-level `derivatives.yaml`, but also writes C++ functions to handle it.

Do we need to go that far?
Maybe, if this loss function becomes public infrastructure in the future and is used by people on different software and hardware architectures.
But for now, we have a way to modify the computation graph at the application layer; we only need to write R code.
The framework designers left us a back door: torch's `autograd_function`.

```{r}
?autograd_function
```
With it, we can customize our forward-propagation and backward-propagation rules.
For forward propagation, there is nothing that needs to change.
For backward propagation, we can reach in, cancel these terms, eliminate the denominator, and rescue it from autograd's rigid rules.

The documentation is clear.
Let us write it now.

```{r}
fused_hellinger_loss <- autograd_function(
  forward = function(ctx, logits, targets) {
    p <- nnf_softmax(logits, dim = -1) # use softmax on the final dimension to convert logits into probabilities

    targets_2d <- targets$unsqueeze(2)
    p_t <- p$gather(dim = 2, index = targets_2d)$squeeze(2)

    loss <- (- torch_sqrt(p_t))$mean() # for multiple batches and samples, take the average

    ctx$save_for_backward(p = p, p_t = p_t, targets = targets) # save intermediate results

    return(loss)
  },

  backward = function(ctx, grad_output) {
    # retrieve what we stored during the forward phase
    saved <- ctx$saved_variable
    p <- saved[[1]]  # the first saved object is p
    p_t <- saved[[2]]  # the second saved object is p_t
    targets <- saved[[3]]  # the third saved object is targets

    batch_size <- p$shape[1]

    sqrt_p_t <- torch_sqrt(p_t)$unsqueeze(2)  # shape: (batch, 1)

    # 1/2 * (pi - delta_ti) * sqrt(pt)
    grad_logits <- 0.5 * p * sqrt_p_t
    grad_true_class <- torch_zeros_like(p)$scatter(
      dim = 2,
      index = targets$unsqueeze(2),
      src = 0.5 * sqrt_p_t
    )
    grad_logits <- grad_logits - grad_true_class
    grad_logits <- grad_logits * grad_output / batch_size

    return(list(logits = grad_logits, targets = NULL))
  }
)
```

I want to explain this code a little more so that the details are clear.
First, this code considers the multi-sample case.
It assumes that the input `logits` has shape `(batch_num, class_num)`, and `targets` is a one-dimensional tensor with shape `(batch_num,)`.
Of course, one could also choose to use `flatten()` to flatten it.

`p <- nnf_softmax(logits, dim = -1)` means applying softmax on the final dimension, i.e. the class dimension.
The shape does not change; it is still `(batch_num, class_num)`.

`targets$unsqueeze(2)` adds one dimension to the one-dimensional tensor, turning it into `(batch_num, 1)`, so that it can work with the later `gather` function.
`p$gather(dim = 2, index = targets_2d)` means: along the second dimension, i.e. the class dimension, pick out the indices in `target_2d`.
This obtains the true-class \(p_t\).
Then `squeeze(2)` changes the shape back to `(batch_num,)`.

`loss <- (- torch_sqrt(p_t))$mean()` averages the loss over the current batch.
Of course, if we consider imbalance, we would not simply average; we would adjust according to the size of each class.
Here I only show the simplest case.
If we really run empirical tests, this needs to be combined with samplers, which is outside the scope of this post.

Finally, `ctx$save_for_backward(p = p, p_t = p_t, targets = targets)` saves intermediate results for use during backward propagation.
The forward pass returns the computed loss.

The backward-propagation part is a bit harder to understand starting from `grad_logits`.
First remember the identity obtained by algebraic transformation:

$$
\frac{1}{2}(p_i-\delta_{ti})\sqrt{p_t}.
$$

The line:

```{r}
grad_logits <- 0.5 * p * sqrt_p_t
```

is part of the final identity:

$$
\frac{1}{2}p_i\sqrt{p_t}.
$$

Here `p` has shape `(batch_num, class_num)`, while `sqrt_p_t` has shape `(batch_num, 1)` because of the previous `unsqueeze`.
This triggers broadcasting of `sqrt_p_t`, so every \(p_i\) is multiplied by \(\sqrt{p_t}\).

For non-true classes, this completes the calculation.
But for the true class, where \(i=t\), we are not done.
Therefore, we create `grad_true_class`.
For non-true classes it is 0; for the true class, we still need to subtract \(\frac{1}{2}\sqrt{p_t}\).
Finally, we divide by `batch_size` because we also averaged the loss.
Do not forget `grad_output`, the upstream gradient, which must be multiplied in according to the chain rule.
The final `grad_logits` still has shape `(batch_num, class_num)`.

```{r}
grad_true_class <- torch_zeros_like(p)$scatter(
  dim = 2,
  index = targets$unsqueeze(2),
  src = 0.5 * sqrt_p_t
)
```

How should we understand this key piece of code?

`torch_zeros_like(p)` creates an all-zero matrix with shape `(batch_num, class_num)`, a canvas.
The `scatter` operation with `dim = 2` tells the sniper rifle: fire along the column, i.e. class, dimension.
`index = targets$unsqueeze(2)` says where the bullet should hit.
`src = 0.5 * sqrt_p_t` says what powder is inside the bullet:
put \((batch_num, 1)\)-shaped \(\frac{1}{2}\sqrt{p_t}\) into it.

Written!

It is worth noting that the final returned `grad_logits` still has shape `(batch_num, class_num)`.
The gradient for every logit of every sample is preserved.

```{r}
logits <- torch_tensor(rbind(
  c(1500.0, 0.0, 0.0), # confidently correct; true class is 1
  c(0.0, 0.0, 1500.0)  # confidently wrong; true class is 2
), requires_grad = TRUE)

targets <- torch_tensor(c(1, 2), dtype = torch_long())

loss <- fused_hellinger_loss(logits, targets)
cat(sprintf("current Batch loss: %.4f\n", as.numeric(loss)))

loss$backward()

cat("\n--- Final returned logits gradient ---")
logits$grad
```
What if we wrap it with the outer square root, \(\sqrt{1-\sqrt{p_t}}\)?

For true-class logits, using some algebraic simplification tricks, namely multiplying numerator and denominator by \(\sqrt{1-\sqrt{p_t}}\), then multiplying both by \(1+\sqrt{p_t}\), gives the gradient:

$$
-\frac{1}{4}\sqrt{1-\sqrt{p_t}}(1+\sqrt{p_t})\sqrt{p_t}.
$$

No troublesome denominator.

But for a non-true class:

$$
\frac{1}{4}
\frac{\sqrt{1-\sqrt{p_t}}(1+\sqrt{p_t})}{1-p_t}
\sqrt{p_t}p_f.
$$

The denominator cannot be eliminated.

The non-true-class case is even funnier:
the denominator is 0 when \(1-p_t=0\), i.e. when \(p_t\) approaches 1, meaning the model is approaching success.
When it is 0, there is no gradient; when \(p_f=0\), there will be `NaN`.

This is a bit hard. I did not think of a solution immediately.
But \(1-p_t\) is exactly the sum of all \(p_f\)'s, isn't it?
Maybe if the non-true classes can be added together, this denominator can be eliminated.
Let us discuss this in the future.

#### 3.3.2 Protect the Square Root

Since `sqrt` is the root of all evil, we need to protect it carefully.

We will write our own `safe_sqrt`.
The goals:

1. Do not ruin forward propagation. If `safe_sqrt` computes \(\sqrt{0}=1\), it breaks the forward pass, even though we will go through this step later.
2. Make backward propagation safe: do not allow 0 to appear in the denominator.

Now we are short-sighted people.
We focus too much on preventing 0 from appearing in the denominator, and ignore whether forward propagation lives or dies.
So when the troublemaker 0 appears, we say: no, you are not allowed to be 0; now you are 1.

```{r}
safe_sqrt_myopic <- function(x){
    x_safe <- torch_where(x > 0, x, x + torch_ones_like(x))
    return(torch_sqrt(x_safe))
}

zero <- torch_tensor(0, dtype = torch_float32(), requires_grad = TRUE)
result <- safe_sqrt_myopic(zero)
result$backward()
cat("result (forward): ", as.numeric(result), "\n")
cat("grad (back): ", as.numeric(zero$grad), "\n")
```
We bent the gradient from `Inf` into 0.5, because we forcibly shifted \(x\) by a constant 1.
Whether the shift is 1, 2, or any other number is a design degree of freedom.

But the forward pass is broken.
The next step is to fix the forward pass.
It only breaks when \(x=0\):

```{r}
safe_sqrt <- function(x){
    x_safe <- torch_where(x > 0, x, x + torch_ones_like(x))
    return(torch_sqrt(x_safe) * torch_where(x > 0,
    torch_ones_like(x), torch_zeros_like(x)))
}

zero <- torch_tensor(0, dtype = torch_float32(), requires_grad = TRUE)
result <- safe_sqrt(zero)
result$backward()
cat("result for 0 (forward): ", as.numeric(result), "\n")
cat("grad for 0 (back): ", as.numeric(zero$grad), "\n")

one <- torch_tensor(1, dtype = torch_float32(), requires_grad = TRUE)
result <- safe_sqrt(one)
result$backward()
cat("result for 1 (forward): ", as.numeric(result), "\n")
cat("grad for 1 (back): ", as.numeric(one$grad), "\n")
```
Fixed.

The advantage of this method is that it is relatively general.
It applies to:

$$
\sqrt{\sum(\text{one-hot}-\sqrt{p_k})^2},
$$

and also to:

$$
\sqrt{1-\sqrt{p_t}}
$$

and:

$$
1-\sqrt{p_t}.
$$

All of them can be handed to autograd without problems.
The concern is that this changes the training dynamics:
where there should have been an infinite gradient at \(x=0\), it now becomes an arbitrarily designed number.
The dynamics change.
But we can design it to shift by a very large number, so that when \(x=0\), or when it underflows to 0, there is still plenty of force.

#### 3.3.3 Other Solutions

I collectively call these epsilon schemes.
These methods are all direct, and their ideas are simple.

For example, `clamp`: since numerical instability appears when \(p\) reaches an extreme value, we prevent it from reaching that extreme by restricting it to \((0+\varepsilon, 1-\varepsilon)\).

Or more directly:

$$
\sqrt{\max(x,\varepsilon)}.
$$

Before entering \(\sqrt{\cdot}\), lift any input smaller than \(\varepsilon\) up to \(\varepsilon\), so \(\sqrt{\cdot}\) never sees a number smaller than \(\varepsilon\).

Another example: protect `sqrt` with epsilon:

$$
x = \sqrt{x+\varepsilon^2}.
$$

These methods are not discussed further here.
They will be discussed in future posts during empirical tests on real datasets of various scales.

## 4. Philosophical Reflection and Preview of the Next Episode

Minimizing \(1-\sqrt{p_t}\) and minimizing \(-\log(p_t)\) are both maximizing \(p_t\), aren't they?
Are they not the same thing?
And even more directly, what about making the loss function \(-p_t\)?

The answer is training dynamics:
how to train more effectively and reach our goal faster.

For example, after gradient analysis, \(-\sqrt{p_t}\) has a vanishing gradient at 0.
When \(p_t=0\), \(-\log(p_t)\) has strong gradient force.

Some people have proposed q-loss, also called generalized cross entropy:

$$
\frac{1-p_t^q}{q},
$$

where \(q\) is between 0 and 1.

Reference:

```bibtex
@misc{zhang2018generalizedcrossentropyloss,
      title={Generalized Cross Entropy Loss for Training Deep Neural Networks with Noisy Labels},
      author={Zhilu Zhang and Mert R. Sabuncu},
      year={2018},
      eprint={1805.07836},
      archivePrefix={arXiv},
      primaryClass={cs.LG},
      url={https://arxiv.org/abs/1805.07836},
}
```

Using L'Hopital's rule to take the gradient, when \(q \rightarrow 0\), the gradient of:

$$
\frac{1-p_t^q}{q}
$$

is:

$$
-p_t^{-1}.
$$

Isn't this exactly the familiar gradient of Cross Entropy?
So the paper says that when \(q\) approaches 0, it becomes CE, more rigorously called Categorical Cross Entropy (CCE).

What is \(q=1/2\)?
Doesn't it look like the simplified Hellinger Distance?

Whether it is a numerical-stability problem or a training-dynamics problem, in the end, we still need many experiments to validate our various schemes and ideas.
Stay tuned for the next post in this series.
