---
title: "Proportional versus Random Ordering: Statistical Analysis and Benchmarking"
description: ""
author: "Wei Cao"
date: "2026-06-11"
categories: []
fig-width: 10
fig-height: 4
fig-dpi: 100
---

```{r}
#| include: false
options(width = 120)
```

The goal of this post is to do some statistical analysis to different ordering strategies
described in Prof. Hocking's blog:
[Creating large imbalanced data benchmarks](https://tdhock.github.io/blog/2025/imbalance-openml/label-order/), and verify our analysis by benchmarking with [atime](https://github.com/tdhock/atime).

## Motivation

When we do subsampling of a dataset, we always want to preserve the original class distribution.
The most natural way is to do random sampling: we take a random sample then do the next without putting it back,
believing the law of large numbers works. That is true, but we can do better.

The aforementioned blog provides proportional ordering with R implementation (we will show later)
and claims that "the proportional ordering is somewhat more stable than the random ordering".

Why, and how much is it more stable?

## Data Preparation

 We download Parquet instead of ARFF files. Please refer to [this article](https://weicaocw.github.io/mlr3hf/entry_tasks/easy/live_report/easy_task_live_report.html#arff-vs.-parquet) for more details about the choice.

```{r }
library(data.table)
library(DBI)
library(atime)
library(ggplot2)

did <- 45570L  # OpenML Higgs (11,000,000 lines)
pq_url <- sprintf(
  "https://data.openml.org/datasets/%04d/%04d/dataset_%d.pq",
  did %/% 10000L, did, did)

cache_dir <- tools::R_user_dir("mlr3torchAUM-blog", "cache")
dir.create(cache_dir, recursive = TRUE, showWarnings = FALSE)
pq_local <- file.path(cache_dir, sprintf("dataset_%d.pq", did))

if (!file.exists(pq_local)) {
  message("download ~892 MB (only first time)...")
  curl::multi_download(pq_url, pq_local, resume = TRUE)
}
sprintf("local parquet: %.0f MB", file.size(pq_local) / 1e6)
```


```{r}
con <- dbConnect(duckdb::duckdb())
label_dt <- setDT(dbGetQuery(
  con, sprintf('SELECT "Target" FROM read_parquet(%s)', dbQuoteString(con, pq_local))))
dbDisconnect(con, shutdown = TRUE)
```

```{r}
y_all <- factor(as.character(label_dt$Target))
counts <- label_dt[, .(count = .N), by = Target][order(-count)]
cls0 <- as.character(counts$Target[1])
p0 <- counts$count[1] / sum(counts$count)
Npop <- length(y_all)
counts[, prop := count/sum(count)][]
```

Two classes: approximately 53% vs. 47%.


## Method / implementation

### Reproduce Hocking's two orderings
```{r}

```

## Conclusions

<!-- 用 bullet 列出本文展示了什么 / 学到了什么，
     可指向后续工作或相关软件包。 -->

## Session info

```{r}
sessionInfo()
```
