---
title: "Proportional versus Random Ordering: Statistical Analysis and Benchmarking"
description: ""
author: "Wei Cao"
date: "2026-06-11"
categories: []
fig-width: 10
fig-height: 4
fig-dpi: 100
---

```{r}
#| include: false
options(width = 120)
```

The goal of this post is to do some statistical analysis to different ordering strategies
described in Prof. Hocking's blog:
[Creating large imbalanced data benchmarks](https://tdhock.github.io/blog/2025/imbalance-openml/label-order/), and verify our analysis by benchmarking with [atime](https://github.com/tdhock/atime).

## Motivation

When we do subsampling of a dataset, we always want to preserve the original class distribution.
The most natural way is to do random sampling: we take a random sample then do the next without putting it back,
believing the law of large numbers works. That is true, but we can do better.

The aforementioned blog provides proportional ordering with R implementation (we will show later)
and claims that "the proportional ordering is somewhat more stable than the random ordering".

Why, and how much is it more stable?

## Data Preparation

 We download Parquet instead of ARFF files. Please refer to [this article](https://weicaocw.github.io/mlr3hf/entry_tasks/easy/live_report/easy_task_live_report.html#arff-vs.-parquet) for more details about the choice.

```{r }
library(data.table)
library(DBI)
library(atime)
library(ggplot2)

did <- 45570L  # OpenML Higgs (11,000,000 lines)
pq_url <- sprintf(
  "https://data.openml.org/datasets/%04d/%04d/dataset_%d.pq",
  did %/% 10000L, did, did)

cache_dir <- tools::R_user_dir("mlr3torchAUM-blog", "cache")
dir.create(cache_dir, recursive = TRUE, showWarnings = FALSE)
pq_local <- file.path(cache_dir, sprintf("dataset_%d.pq", did))

if (!file.exists(pq_local)) {
  message("download ~892 MB (only first time)...")
  curl::multi_download(pq_url, pq_local, resume = TRUE)
}
sprintf("local parquet: %.0f MB", file.size(pq_local) / 1e6)
```


```{r}
con <- dbConnect(duckdb::duckdb())
label_dt <- setDT(dbGetQuery(
  con, sprintf('SELECT "Target" FROM read_parquet(%s)', dbQuoteString(con, pq_local))))
dbDisconnect(con, shutdown = TRUE)
```

```{r}
y_all <- factor(as.character(label_dt$Target))
counts <- label_dt[, .(count = .N), by = Target][order(-count)]
cls0 <- as.character(counts$Target[1])
p0 <- counts$count[1] / sum(counts$count)
Npop <- length(y_all)
counts[, prop := count/sum(count)][]
```

Two classes: approximately 53% vs. 47%.


## Method / implementation

### Reproduce Hocking's two orderings (proportional vs. random)
```{r}
make_orders <- function(y) {
  n <- length(y)
  rand_ord <- sample(n)
  prop_ord <- data.table(y = y[rand_ord])[
    , prop_y := seq(0, 1, l = .N), by = y][, order(prop_y)]
  list(random = rand_ord, proportional = rand_ord[prop_ord])
}

set.seed(1)
ord_list <- make_orders(y_all)
rows <- list()
for (ord_name in names(ord_list)) {
  y_ord <- y_all[ord_list[[ord_name]]]
  for (prop_data in c(0.01, 0.1, 1)) {
    n <- as.integer(Npop * prop_data)
    tab <- data.table(y = y_ord[1:n])[, .(prop_y = .N / n), by = y]
    rows[[paste(ord_name, prop_data)]] <- data.table(ord_name, prop_data, tab)
  }
}
dcast(rbindlist(rows), ord_name + prop_data ~ y, value.var = "prop_y")
```

 We reproduced the data table in Hocking's blog. Here Hocking claims that "the proportional ordering is somewhat more stable than the random ordering".


### Statistical Analysis

We view random sampling as a random sampling without replacement, and a single sample is a [Bernoulli](https://en.wikipedia.org/wiki/Bernoulli_distribution) random variable with a variance of $p(1-p)$. With the assumption of independence, the variance of the sample mean is $\frac{p(1-p)}{n}$, therefore the standard deviation is $\sqrt{\frac{p(1-p)}{n}}$.

(TODO) The process is in fact a random sampling with replacement...

As for the proportional ordering: $\frac{1}{n}$ is a constant.

The result will be clear in a log-log plot.


### Use atime to benchmark
```{r}
set.seed(1)
prop_ord_full <- make_orders(y_all)$proportional
is0_ord <- as.integer(y_all[prop_ord_full] == cls0)
cum_p0 <- cumsum(is0_ord) / seq_along(is0_ord)

timing <- atime::atime(
  N=as.integer(10^seq(2, 7, by = 0.25)),
  setup={y <- y_all[1:N] },
  random={sample(length(y)) },
  proportional={ ro <- sample(length(y)); data.table(y = y[ro])[, p := seq(0, 1, l = .N), by = y][, order(p)] },
  seconds.limit=0.1, times=10)

best <- atime::references_best(timing)
plot(best) + ggplot2::ggtitle("Construction time of two orderings vs N (atime + references_best fit Big O)")
```


```{r}
# Delay: median time at a specific N
timing$measurements[N == max(N), .(expr.name, seconds = signif(median, 3))]

# Throughput: N that can be processed within 0.1 seconds (log-log interpolation to solve T(N)=0.1)
predict(best, seconds = 0.1)
```

```{r}
## Let atime use RMSE as the measured unit. random use resampling to estimate RMSE;
## proportional use the deterministic cum_p0 table (O(1)).
rmse_unit <- atime::atime(
  N = as.integer(10^seq(2, 6, by = 0.25)),
  setup = {},
  random       = { ph <- replicate(100, mean(sample(y_all, N) == cls0)); sqrt(mean((ph - p0)^2)) },
  proportional = abs(cum_p0[N] - p0),
  result = function(x) data.frame(rmse = x),
  times = 5, seconds.limit = Inf)

rmse_dt <- rmse_unit$measurements[rmse > 0]
rmse_slopes <- rmse_dt[, .(slope = coef(lm(log10(rmse) ~ log10(N)))[2]), by = expr.name]
print(rmse_slopes)  # should get random ≈ -0.5, proportional ≈ -1

ggplot(rmse_dt, aes(N, rmse, color = expr.name)) +
  geom_point(size = 2) + geom_line() +
  scale_x_log10() + scale_y_log10() +
  labs(title = "atime uses result= to measure not time, but statistical convergence rate",
       y = "RMSE (custom unit, log)", x = "N (log)")
```

## Conclusions

TODO


## Session info

```{r}
sessionInfo()
```
